Let's talk about Thurstone \& Co.: An information-theoretical model for comparative judgments, and its statistical translation

### 2024-11-15 _tex/index.tex
Built site for gh-pages

In \emph{comparative judgment} (CJ) studies, judges assess a specific
trait or attribute across various stimuli by performing pairwise
comparisons \citep{Thurstone_1927, Pollitt_2004, Pollitt_2012a}. Each
comparison produces a dichotomous outcome, indicating which stimulus is
perceived to exhibit a higher trait level. For example, when assessing
text quality, judges compare pairs of written texts (the stimuli) to
determine the relative quality each text exhibit (the trait)
\citep{Laming_2004, Pollitt_2012b, Whitehouse_2012, vanDaal_et_al_2016, Lesterhuis_2018, Coertjens_et_al_2017, Goossens_et_al_2018, Bouwer_et_al_2023}.

@@

Nevertheless, despite the increasing number of CJ studies, unsystematic
and fragmented research approaches have left several critical issues
unaddressed. This research primarily focuses on three: the over-reliance
on Thurstone's Case V assumptions in the statistical analysis of CJ
data, the apparent disconnect between CJ's trait measurement and
hypothesis testing, and the unclear role and influence of comparison
algorithms on the method's reliability and validity. The following
sections will discuss each of these issues in detail, followed by the
introduction of a theoretical model and its statistical translation,
which aims to address all three concerns simultaneously.

@@

In its most general form, Thurstone's theory
\citeyearpar{Thurstone_1927} posits that when a judge compares two
stimuli, the resulting dichotomous outcome depends on two main factors:
the discriminal process and the law of comparative judgment. The
\emph{discriminal process} refers to the psychological effect each
stimulus has on the judges, or more simply stated, the judges'
perception of the trait level of each stimulus. Thurstone assumes that
the discriminal process for each stimulus follows a Normal distribution.
In this distribution, the mode (mean), known as the \emph{modal
discriminal process}, represents the position of the stimulus on the
trait continuum, while the dispersion, known as the \emph{discriminal
dispersion}, reflects variability in the perceived trait level of the
stimulus. Figure~\ref{fig-discriminal_process} shows a visual depiction
of the discriminal process of two stimuli (objects).

@@

\caption{\label{fig-discriminal_process}Discriminal processes of two
stimuli (objects). Extracted from \citet[pp.~249]{Bramley_2008}}

@@

However, since the discriminal mode and dispersion of a single stimulus
are not directly observable without comparing it to other stimuli, the
\emph{law of comparative judgment} becomes essential. This law asserts
that when assessing a specific trait by comparing two stimuli, the
stimulus positioned further along the continuum is perceived as having a
higher level of that trait. Thus, the observed dichotomous outcome is
determined by the distribution of the difference between the discriminal
processes of the two stimuli, called the \emph{discriminal difference}.
Figure~\ref{fig-comparative_judgment} shows a visual depiction of the
discriminal difference of two stimuli (objects).

@@

\caption{\label{fig-comparative_judgment}Discriminal difference between
two stimuli (objects). Extracted from \citet[pp.~251]{Bramley_2008}}

@@

Thurstone's general form of the theory primarily applies to pairwise
comparisons of stimuli made by a single judge
\citep[pp.~267]{Thurstone_1927}. Therefore, to enhance the theory's
practical applicability, he developed five distinct cases, each
incorporating a series of increasingly simplifying assumptions.
Figure~\ref{fig-thurstone_cases} outlines each case along with their key
assumptions. For a detailed explanation of this theoretical progression,
see \citet{Thurstone_1927} and \citet[pp.~248-253]{Bramley_2008}.

@@

Surprisingly, Case V is the most commonly used case in the CJ
literature, despite incorporating the largest number of simplifying
assumptions \citep[pp.~253]{Bramley_2008}. This widespread adoption is
primarily due to the Bradley-Terry-Luce (BTL) model
\citep{Bradley_et_al_1952, Luce_1959}, which offers a simpler
statistical representation of the case. The BTL model closely aligns
with Case V's assumptions, with one key difference: the distribution of
the discriminal process. While Case V assumes a Normal distribution, the
BTL model uses the more mathematically tractable logistic distribution
\citep[pp.~254]{Bramley_2008} (see Figure~\ref{fig-thurstone_cases}).
However, this adjustment does not significantly affect the model's
estimation or interpretation, as the logistic and normal distributions
are related to each other by a factor of approximately \(1.7\)
\citep[pp.~16]{vanderLinden_et_al_2017_I}.

However, because Case V was originally developed to provide a ``rather
coarse scaling'' of traits \citep[pp.~269]{Thurstone_1927}, prioritizing
statistical ease of use over precision in trait measurement, two of its
assumptions may be causing more problems than expected: the assumption
of zero correlation between the stimuli, and the assumption of equal
dispersion between stimuli.

\subsection{The disconnect between structural and measurement
models}\label{the-disconnect-between-structural-and-measurement-models}

As outlined in the previous section, the BTL model typically functions
as the measurement model in a CJ study
\citep{Andrich_1978, Bramley_2008}. A measurement model specifies how
manifest variables contribute to the estimation of latent variables
\citep{Everitt_et_al_2010}. For example, when evaluating text quality,
researchers use the BTL model to process the dichotomous outcomes
resulting from the pairwise comparisons (the manifest variables) to
estimate scores that reflect the underlying quality level of the texts
(the latent variable)
\citep{Laming_2004, Pollitt_2012b, Whitehouse_2012, vanDaal_et_al_2016, Lesterhuis_2018, Coertjens_et_al_2017, Goossens_et_al_2018, Bouwer_et_al_2023}.

>>>>

In \emph{comparative judgment} (CJ) studies, judges assess a specific
trait or attribute across various stimuli by performing pairwise
comparisons \citep{Thurstone_1927a, Thurstone_1927b}. Each comparison
produces a dichotomous outcome, indicating which stimulus is perceived
to exhibit a higher trait level. For example, when assessing text
quality, judges compare pairs of written texts (the stimuli) to
determine the relative quality each text exhibit (the trait)
\citep{Laming_2004, Pollitt_2012b, Whitehouse_2012, vanDaal_et_al_2016, Lesterhuis_2018, Coertjens_et_al_2017, Goossens_et_al_2018, Bouwer_et_al_2023}.

@@

Nevertheless, despite the increasing number of CJ studies, unsystematic
and fragmented research approaches have left several critical issues
unaddressed. This research primarily focuses on three: the over-reliance
on Thurstone's Case V assumptions in the statistical analysis of CJ
data, the apparent disconnect between CJ's trait measurement and
hypothesis testing, and the unclear role of comparison algorithms on the
method's reliability and validity. The following sections will discuss
each of these issues in detail, followed by the introduction of a
theoretical model and its statistical translation, which aims to address
all three concerns simultaneously.

@@

In its most general form, Thurstone's theory
\citeyearpar{Thurstone_1927b} posits that the dichotomous outcome
resulting from comparing two stimuli is determined by two factors: the
discriminal process of each stimulus and their discriminal difference.
The \emph{discriminal process} refers to the psychological effect each
stimulus has on the judges, or more simply stated, the judges'
perception of the trait level of each stimulus. Thurstone assumes that
the discriminal process for each stimulus follows a Normal distribution.
In this distribution, the mode (mean), known as the \emph{modal
discriminal process}, represents the position of the stimulus on the
trait continuum, while the dispersion, known as the \emph{discriminal
dispersion}, reflects variability in the perceived trait level of the
stimulus. Figure~\ref{fig-discriminal_process} shows example
distributions of discriminal process for two stimuli (objects).

@@

\caption{\label{fig-discriminal_process}Example distributions of
discriminal processes for two stimuli (objects). Extracted from
\citet[pp.~249]{Bramley_2008}.}

@@

However, since the discriminal mode and dispersion of a single stimulus
are not directly observable without comparing it to other stimuli, the
\emph{law of comparative judgment} becomes essential. This law asserts
that when assessing a specific trait by comparing two stimuli, the
stimulus positioned further along the continuum is perceived as having a
higher level of that trait. Thus, the observed dichotomous outcome is
determined by the distribution of the difference between the discriminal
processes of the two stimuli, called the \emph{discriminal difference}.
Figure~\ref{fig-comparative_judgment} shows an example distribution of
the discriminal difference for two stimuli (objects).

@@

\caption{\label{fig-comparative_judgment}Distribution of the discriminal
difference between two stimuli (objects). Extracted from
\citet[pp.~251]{Bramley_2008}.}

@@

Importantly, the theory's general form primarily addresses pairwise
comparisons of stimuli made by a single judge
\citep[pp.~267]{Thurstone_1927b}. Consequently, to enhance its practical
applicability, Thurstone introduced five distinct cases, each defined by
progressively simplifying assumptions. Figure~\ref{fig-thurstone_cases}
summarizes these cases, focusing on key assumptions such as the
distribution of discriminal processes, the similarity of discriminal
dispersion across stimuli, the correlation between stimuli, and which
judges perform the comparisons. For a comprehensive discussion of this
progression, refer to \citet{Thurstone_1927b} and
\citet[pp.~248-253]{Bramley_2008}.

@@

Surprisingly, despite its reliance on the largest number of simplifying
assumptions \citep[pp.~253]{Bramley_2008}, Case V remains the most
widely used case in the CJ literature. This popularity is largely due to
its simplified statistical representation in the Bradley-Terry-Luce
(BTL) model \citep{Bradley_et_al_1952, Luce_1959}. The BTL model mirrors
Case V's assumptions, with one key difference: while Case V assumes a
Normal distribution for the discriminal processes, the BTL model uses
the more mathematically tractable logistic distribution
\citep[pp.~254]{Bramley_2008} (see Figure~\ref{fig-thurstone_cases}).
This substitution has little effect on the model's estimation or
interpretation, as the Normal and logistic distributions differ by a
scaling factor of approximately \(1.7\)
\citep[pp.~16]{vanderLinden_et_al_2017_I} (refer to
Figure~\ref{fig-logistic_vs_normal}).

\begin{minipage}{0.50\linewidth}

\includegraphics[width=0.95\linewidth,height=\textheight,keepaspectratio]{./images/figures/density.png}

\subcaption{\label{fig-density}Probability density}

\end{minipage}%
%
\begin{minipage}{0.50\linewidth}

\includegraphics[width=0.95\linewidth,height=\textheight,keepaspectratio]{./images/figures/cumulative.png}

\subcaption{\label{fig-cumulative}Cummulative probability}

\end{minipage}%

\caption{\label{fig-logistic_vs_normal}Probability density and
cumulative probability of the logistic and Normal distributions.
Extracted from \citet[pp.~254-255]{Bramley_2008}.}

However, Case V (and its BTL model counterpart) was originally developed
to provide a ``rather coarse scaling'' of traits
\citep[pp.~269]{Thurstone_1927b}, prioritizing statistical simplicity
over precision in trait measurement. As a result, its assumptions may
not be suitable for applications beyond the psychophysical contexts,
which typically involve simple, homogeneous stimuli. Thurstone himself
cautioned that its use ``should not be made without (an) experimental
test'' \citep[pp.~270]{Thurstone_1927b}, further acknowledging that some
assumptions could become problematic with less homogeneous stimuli, such
as handwriting or English compositions \citep[pp.~374]{Thurstone_1927a}.
Consequently, given that current CJ applications often involve more
complex, less homogeneous stimuli, two assumptions of Case V may not
always hold theoretical or in practice: the assumption of zero
correlation and equal dispersion between stimuli.

{Regarding the assumption of \emph{zero correlation between stimuli}
\ldots{}}

\subsection{The disconnect between trait measurement and hypothesis
testing}\label{the-disconnect-between-trait-measurement-and-hypothesis-testing}

Building on the previous section, it is evident that the BTL model
typically functions as the measurement model for the trait under
assessment \citep{Andrich_1978, Bramley_2008}. A measurement model
specifies how manifest variables contribute to the estimation of latent
variables \citep{Everitt_et_al_2010}. For example, when evaluating text
quality, researchers use the BTL model to process the dichotomous
outcomes resulting from the pairwise comparisons (the manifest
variables) to estimate scores that reflect the underlying quality level
of the texts (the latent variable)
\citep{Laming_2004, Pollitt_2012b, Whitehouse_2012, vanDaal_et_al_2016, Lesterhuis_2018, Coertjens_et_al_2017, Goossens_et_al_2018, Bouwer_et_al_2023}.
